Palmer amaranth (\emph{Amaranthus palmeri}) adaptation to US Midwest agroecosystems

### 2021-09-17 Manuscript/Manuscript.tex
as

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is one of the most troublesome weed species in the United States. Palmer amaranth is endemic to Southern United States but its range is expanding northward. Palmer amaranth dispersal warrants studies assessing species adaptation into new geographies. A study was conducted to investigate morphology, flowering and sex dimorphism from cohorts of Palmer amaranth growing under corn, soybean, and bareground across five locations of US Midwest. Results demonstrated that first cohort of Palmer amaranth, established in June, produced 42\% more biomass than plants from second cohort (established in July). The first Palmer amaranth cohort produced 75.5 g plant\textsuperscript{-1} in bareground, 28.3 g plant\textsuperscript{-1} in soybean and 16.3 g plant\textsuperscript{-1} in corn, whereas the second Palmer amaranth cohort produced 62.6, 6.3, and 1.4 g plant\textsuperscript{-1} in bareground, soybean and corn, respectively. Palmer amaranth height was more impacted when growing in corn. Palmer amaranth plants averaged 85.2 cm tall in the first cohort but 38.2 cm tall in the second cohort in corn. Moreover, Palmer amaranth flowering window shifted according to crop and cohort timings. Palmer amaranth growing in intense competition, such as under low light in corn, resulted in the longest flowering window. Also, Palmer amaranth sex dimorphism was slightly influenced by day of year, weight and height. The model estimated that probability of being a female plant increased as biomass and height increased. Our results showed the fast adaptation and plasticity of Palmer amaranth to grow and adapt to cropping systems from the US Midwest. Palmer amaranth is likely to continue its expansion northward. Therefore, preventing plant dispersal into new habitats is the most effective management strategy. Reactive management to reduce Palmer amaranth impact on cropping systems should encompass diversity of tactics that minimize the species ability to establish into cropping systems, including crop rotation (beyond corn and soybean), early/late crop planting, row spacing, cover crops, and effective chemical control programs.

@@

Palmer amaranth is a fast growing summer annual forb indigenous to the
Sonoran Desert (Sauer, 1957). The species would eventually emerge as a
threat to US agriculture in the 1990s. Palmer amaranth weediness is
likely a result of human-assisted selection in combination with plant
biology. Farm mechanization, adoption of conservation agriculture (e.g.,
no-till), and intensive use of herbicides for weed management are the
main human-mediated selection of Palmer amaranth into cropping systems
(Ward et al., 2013). On the other hand, Palmer amaranth is a prolific
seed producer with a C4 photosynthetic apparatus (Wang et al., 1992).
With a dioecy nature, Palmer amaranth male and female plants are
obligate outcrosser species, increasing the chances of exchanging
adaptive traits among plants (Jhala et al., 2021; Oliveira et al.,
2018). Also, Palmer amaranth small seeds (e.g, 1 mm) tend to thrive in
no-tillage systems (Price et al., 2011), and spread across locations
through farm equipment (Sauer, 1972), seed mixes (Hartzler and Anderson,
2016), wildlife (Farmer et al., 2017), etc. The dispersal capacity of
Palmer amaranth makes the species one of the most successful cases of
weed adaptation to current cropping systems.

Light and temperature are the main environmental requirements for Palmer
amaranth growth and development (Jha et al., 2010). Palmer amaranth is
reported with an extended germination period (Ward et al., 2013).
Germination of Palmer amaranth was triggered by 18 C soil temperature at
5 cm depth (Keeley et al., 1987), and optimal germination and biomass
production occurred at 35/30 C day and night temperatures (Guo and
Al-Khatib, 2003). In addition, Palmer amaranth establishment is
human-mediated by tillage timings and preemergence-applied herbicides
(Chahal et al., 2021), which can result in weed germination shifts
(Sbatella and Wilson, 2010). Also, Palmer amaranth has shown to grow and
develop under reduced soil water content. In continuous water stress,
Palmer amaranth survived and produced at least 14000 seeds
plant\textsuperscript{-1} (Chahal et al., 2018). Seeds from Palmer
amaranth growing with limited water conditions were heavier, less
dormant, and prompt for germination (Matzrafi et al., 2021). Growing
conditions and management practices also influence Palmer amaranth sex
dimorphism and flowering pattern (Korres et al., 2017; Rumpa et al.,
2019). Therefore, Palmer amaranth has shown plasticity to evolve and
fast adapt under the current agroecosystem conditions. Further scenarios
show that global temperature warming can impact agriculture, and promote
niches for Palmer amaranth invasion/adaptation into new environments.
Currently, it is estimated that the greatest climatic risk of Palmer
amaranth establishment are agronomic crops in Australia and Sub-Sahara
Africa (Kistner and Hatfield, 2018). Temperature is a key factor
limiting Palmer amaranth expansion to cooler geographies (Briscoe
Runquist et al., 2019); however, under future climate change Palmer
amaranth is likely to expand northward into Canada and Northern Europe
(Kistner and Hatfield, 2018; Briscoe Runquist et al., 2019).

Palmer amaranth is already found in agronomic crops of South America
(Larran et al., 2017; Küpper et al., 2017) and Southern Europe (Milani
et al., 2021). In the United States, Palmer amaranth is well established
in the Cotton Belt (Garetson et al., 2019; Bagavathiannan and
Norsworthy, 2016) in the southern United States but its range is
expanding northward. For example, herbicide resistant Palmer amaranth is
widespread in Nebraska (Oliveira et al., 2021). There are some reported
cases of Palmer amaranth in Michigan (Kohrt et al., 2017) and
Connecticut (Aulakh et al., 2021). Also, it is estimated that Palmer
amaranth can cause high damage to soybean fields in Illinois (Davis et
al., 2015), which is concerning as soybean along with corn make most of
US Midwest agronomic hectares. In Iowa, a study showed that Palmer
amaranth is still not well adapted compared to waterhemp
(\emph{Amaranthus tuberculatus}) (Baker, 2021). Waterhemp, an
\emph{Amaranthus} species, is the most troublesome species in the US
Midwest (Tranel et al., 2011). Nonetheless, invasion and successful
eradication of Palmer amaranth is documented in Minnesota (Yu et al.,
2021). Palmer amaranth infestations have not been detected in Canada;
however, Palmer amaranth seeds were detected in sweet potato slips in
the country (Page et al., 2021). Palmer amaranth is still not as well
adapted and established to Northern as it is in the Southern North
America. Therefore, Palmer amaranth range of expansion into new habitats
can increase. It seems certain the need to manage new Palmer amaranth
infestations in agronomic crops throughout northern United States in the
near future. Strategies on Palmer amaranth management should encompass
the agroecosystem level but not only attempts to eradicate the weed.
Most tactics to manage Palmer amaranth are based on technology fixes
(Scott, 2011), which are short-term (e.g., herbicide and/or tillage)
rather than long-term weed management. Palmer amaranth management should
be built on minimizing the species ability to adapt, grow and develop
into agroecossystems.

In the southeastern US, early growing Palmer amaranth is well known to
have a higher impact on cotton yields compared to late established
plants (MacRae et al., 2013). In the northern states, Palmer amaranth
impact on the agroecosystem is recent. Studies investigating Palmer
amaranth in those locations are limited due to the plant classification
as noxious weed species in some northern states (Yu et al., 2021).
Nonetheless, the continuous Palmer amaranth dispersal and potential
establishment across the northern United States is concerning and
warrants investigations on species morphology in such environments.
Understanding Palmer amaranth biology and growing strategies under
different agroecosystems can enhance our knowledge on species adaptation
and management practices. It can also aid in designing proactive and
ecological tactics to limit the species range expansion, reduce its
negative impact, and developing resilient and sustainable farming
systems (MacLaren et al., 2020). Therefore, the objective of this study
was to investigate the flowering pattern, sex dimorphism, biomass
production, and height of Palmer amaranth cohorts growing under corn,
soybean and bareground across five locations in the United States
Midwest.

@@

\caption{\label{tab:unnamed-chunk-2}Field study attributes }
\centering
\resizebox{\linewidth}{!}{
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lllllll}
\toprule
Attributes &  & Arlington, WI & Clay Center, NE & Grant, NE & Lincoln, NE & Macomb, IL\\
\midrule
Bareground &  &  &  &  &  & \\

 & Weed control & glyphosate & saflufenacil + imazethapyr + pyroxasulfone &  &  & \\

@@

 & Hybrid & NK0142 3120-EZ1 & DKC60-67 &  &  & \\

 & Seeding rate & 88956 & 86487 &  &  & \\

 & Weed control & glyphosate / S-metolachor & S-metolachlor + trazine + mesotrione, + bicyclopyrone &  &  & \\

 & Stage at 1 cohort & V2-3 &  &  &  & \\

 & Stage at 2 cohort & V6-7 &  &  &  & \\

 & Planting day & April 30, 2018 / May 5, 2019 & May 10, 2018/19 &  &  & \\

\multirow{-7}{*}{\raggedright\arraybackslash } & Fertilization & N (46-0-0) at 157 kg ha-1 &  &  &  & \\

@@

 & Variety & DSR-1950 & AG21X8 &  &  & \\

 & Seeding rate & 296400 & 321237 &  &  & \\

 & Weed control & glyphosate / S-metolachor & saflufenacil + imazethapyr + pyroxasulfone &  &  & \\

 & Stage at 1 cohort & V1-2 &  &  &  & \\

 & Stage at 2 cohort & V5-6 &  &  &  & \\

\multirow{-6}{*}{\raggedright\arraybackslash } & Planting day & May 5, 2018 / May 10, 2009 & May 14, 2018/19 &  &  & \\

@@

 & Type & Plano-silt-loam & Crete Silt Loam &  &  & \\

 & Ratio (sand-clay-silt) & 10-64-26 & 17-58-25 &  &  & \\

 & pH & 6.6 & 6.5 &  &  & \\

\multirow{-4}{*}{\raggedright\arraybackslash } & Organic matter (\%) & 3.5 & 3 &  &  & \\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{a} glyphoste, 840 g ae ha \textsuperscript{b} S-metolachor, 1324 g ai ha; \textsuperscript{c} S-metolachlor + trazine + mesotrione, + bicyclopyrone, 2409 g ai ha; \textsuperscript{d} saflufenacil + imazethapyr + pyroxasulfone, 215 g ai ha}\\
\end{tabular}}
\end{table}

The field experimental units were three adjacent 9.1 m wide (12 rows at
76.2 cm row spacing) by 10.7 m long. The experimental design were
arranged in factorial design with three crops, two transplanting times
simulating two cohorts, repeated across five locations. Each filed
experimental unit was planted with corn, soybean, or kept under
bareground. The two transplant timings were June 1 (first cohort) and
July 1 (second cohort). Palmer amaranth seedlings (potting mix + two
seedlings) were transplanted (6 cm deep and 8 cm wide). Forty-eight
plants were equidistantly placed (0.76 m apart) between rows within each
crop. After a week, one plant was eliminated and one was kept, resulting
in 24 plants per experimental unit and transplanting time (Figure 2).
When needed, Palmer amaranth plants were supplied with water during the
first week after transplanting to assure seedling survival.

After transplanting, Palmer amaranth flowering was monitored until the
end of the study. When a plant flowered, the day was recorded, plant sex
was identified (male or female), plant height was measured from soil
surface to the top of plant. Also, aboveground plant biomass was
harvested near soil surface and oven dried at 65 C until reaching
constant weight before weigh (g plant\textsuperscript{-1}) was recorded.

@@

A binary logistic regression was fitted to Palmer amaranth sex
dimorphism. Binary logistic regression is used for predicting binary
classes (Bangdiwala, 2018), such as the probability of a plant being
female in a dioecious species. Prior to the analysis, missing values
were removed from the dataset. Also, data from Grant was not used in
this analysis due to the uniform plant harvesting at that location. The
complete dataset was splitted into 80\% train and 20\% test data. The
80\% train is used for the model training and the 20\% test is used for
checking model performance on unseen dataset. With 80\% dataset, a
generalized linear model (base R \emph{glm} function) was fitted to
binary response variable, the probability of being female (0 to male and
1 to female). The independent variables were day of year harvest,
height, weight, and crop (without interaction). The model family was
binomial with a logit function. The model fit was assessed through
pseudo R-squared values (McFadden, Cox and Snell, Cragg and Uhler) and
likelihood ratio using \emph{nagelkerke} function from ``rcompanion''
package (Mangiafico, 2021). The marginal effects computation was
performed with Average Marginal Effects (AMEs) at every observed value
of x and average across the results (Leeper, 2017) using \emph{margins}
function from ``margins'' package (Leeper et al., 2021). The 20\% test
data was predicted using the \emph{predict} function with a cutoff
estimation for male or female using \emph{performance} function from
ROCR package (Sing et al., 2005). The model quality prediction from the
classification algorithm was measured with precision (\emph{precision}
function), recall (\emph{recall} function) and F1-score (\emph{f\_meas}
function) using the ``yardstick'' package (Kuhn et al., 2021). The
precision determines the accuracy of positive predictions (female
plants), recall determines the fraction of positives that were correctly
identified, and F1-score is a weighted harmonic mean of precision and
recall with the best score of 1 and the worst score of 0 (Raoniar,
2021). F1-score conveys the balance between the precision and the recall
(Yacouby and Axman, 2020). The area under the receiver operating curve
(AUC-ROC) was also estimated with performance function using the true
positive and false positive rates. Higher the AUC, better the model is
at distinguishing between female and male Palmer amaranth.

@@

\hypertarget{palmer-amaranth-sex-dimorphism}{%
\subsection*{Palmer amaranth sex
dimorphism}\label{palmer-amaranth-sex-dimorphism}}
\addcontentsline{toc}{subsection}{Palmer amaranth sex dimorphism}

@@

\caption{\label{tab:unnamed-chunk-3}Average marginal effects of Palmer amaranth sex dimorphism logistic model. Factor pararemter values (crop and bareground) is shown compared to soybean.}
\centering
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lrrrrrr}
\toprule
Term & AME & SE & Lower & Upper & Z-score & P-value\\
\midrule
crop\_bareground & -0.048 & 0.054 & -0.154 & 0.059 & -0.876 & 0.381\\

@@

Our study showed that Palmer amaranth biomass, height, flowering window
and sex varied within crops and cohort timings. In general, first cohort
of Palmer amaranth plants were heavier and taller when compared to the
second cohort. At first cohort, resources (e.g., soil nutrients) and
conditions (e.g., light) were more timely available for the species.
High biomass and taller Palmer amaranth plants are likely a weed
strategy to compete for light in between crop rows in absence of canopy.
In such conditions, Palmer amaranth showed an extraordinary plasticity
to adapt upon the agroecosystem. This is evident when comparing Palmer
amaranth canopy shape. The Palmer amaranth competition (e.g., light)
strategy was to mimic the crop grow and development (Figure 6). These
results suggests that Palmer amaranth can fast evolve life-history
traits to adapt into cropping systems and cultural practices, which was
also showed in a study varying nitrogen fertilization (Bravo et al.,
2018). Our results highlight the Palmer amaranth as a threat to field
crops as breeding more competitive crop varieties is likely to select
more competitive weed biotypes (Bravo et al., 2017).

Palmer amaranth grow and development in second cohort was limited due to
the crop competitive ability at advanced development stages. Plants were
transplanted at greater crop height and width, which reduced Palmer
amaranth competitiveness. As a result, Palmer amaranth height and
biomass was lower compared to its first cohort. Moreover, Palmer
amaranth growing without crop competition produced the highest amounts
of biomass. The Palmer amaranth strategy in bareground was to invest
biomass in growing plant width and height. Nonetheless, Palmer amaranth
produced 17\% less biomass in second cohort compared to first cohort
timing. In a bareground study, early emerged Palmer amaranth without
competition was 50\% taller than late emerged plants (Webster and Grey,
2015). These results suggest that crop competition is not the only
factor limiting late Palmer amaranth establishment. The limited growth
of Palmer amaranth at second cohort is likely a reduced plant response
to day length, light availability and thermal units (e.g, growing degree
days). The \emph{Amaranthus} species are sensitive to photoperiod (Wu
and Owen, 2014). It is hypothesize that reduced day length contributed
to smaller plants at second cohort regardless the crop. A study in North
Carolina and Illinois predicted that less than 10\% Palmer amaranth
seedlings emergence occurred after June (Piskackova et al., 2021). In
addition, Palmer amaranth negative impact on soybean (Korres et al.,
2020) and cotton (Webster and Grey, 2015) yields was higher when plants
were established near to crop planting. Therefore, early management is a
key strategy to minimize the damaging impact of Palmer amaranth to US
Midwest cropping systems.

>>>>

Palmer amaranth (\emph{Amaranthus palmeri} S. Watson) is one of the most troublesome agronomic weed species in the United States. Palmer amaranth is endemic to Southern United States but its range is expanding northward. Palmer amaranth dispersal warrants studies assessing species adaptation into new geographies. A study was conducted to investigate morphology, flowering and gender from cohorts of Palmer amaranth growing under corn, soybean, and bareground across five locations of US Midwest. Results demonstrated that the first cohort of Palmer amaranth, established in June, produced 42\% more biomass than plants from the second cohort (established in July). The first Palmer amaranth cohort produced 75.5 g plant\textsuperscript{-1} in bareground, 28.3 g plant\textsuperscript{-1} in soybean and 16.3 g plant\textsuperscript{-1} in corn, whereas the second Palmer amaranth cohort produced 62.6, 6.3, and 1.4 g plant\textsuperscript{-1} in bareground, soybean and corn, respectively. Palmer amaranth height was more impacted when growing in corn. Palmer amaranth plants averaged 85.2 cm tall in the first cohort but 38.2 cm tall in the second cohort in corn. Moreover, Palmer amaranth flowering window shifted according to crop and cohort timings. Palmer amaranth growing in intense competition, such as under low light in corn, resulted in the longest flowering window. Also, Palmer amaranth gender was slightly influenced by day of year, weight and height. The model estimated that probability of being a female plant increased as biomass and height increased. Our results showed the fast adaptation and plasticity of Palmer amaranth to grow and adapt to cropping systems from the US Midwest. Palmer amaranth is likely to continue its expansion northward. Therefore, preventing plant dispersal into new habitats is the most effective management strategy. Reactive management to reduce Palmer amaranth impact on cropping systems should encompass diversity of tactics that minimize the species ability to establish into cropping systems, including crop rotation (beyond corn and soybean), early/late crop planting, row spacing, cover crops, and effective chemical control programs.

@@

Palmer amaranth is a fast growing summer annual forb indigenous to the
Sonoran Desert (Sauer, 1957). The species would eventually emerge as a
threat to US agriculture in the 1990s. Palmer amaranth weediness is
likely a result of human-assisted selection in combination with plant
biology. Farm mechanization, adoption of conservation agriculture (e.g.,
no-till), and intensive use of herbicides for weed management are the
main human-mediated selections of Palmer amaranth into cropping systems
(Ward et al., 2013). On the other hand, Palmer amaranth is a prolific
seed producer with a C4 photosynthetic apparatus (Wang et al., 1992).
With a dioecious nature, Palmer amaranth male and female plants are
obligate outcrosser species, increasing the chances of exchanging
adaptive traits among plants (Jhala et al., 2021; Oliveira et al.,
2018). Also, Palmer amaranth's small seeds (e.g, 1 mm) tend to thrive in
no-tillage systems (Price et al., 2011), and spread across locations
through farm equipment (Sauer, 1972), seed mixes (Hartzler and Anderson,
2016), wildlife (Farmer et al., 2017), etc. The dispersal capacity of
Palmer amaranth makes the species one of the most successful cases of
weed adaptation to current cropping systems.

Light and temperature are the main environmental requirements for Palmer
amaranth growth and development (Jha et al., 2010). Palmer amaranth is
reported with an extended germination period (Ward et al., 2013).
Germination of Palmer amaranth was triggered by 18 C soil temperature at
5 cm depth (Keeley et al., 1987), and optimal germination and biomass
production occurred at 35/30 C day and night temperatures (Guo and
Al-Khatib, 2003). In addition, Palmer amaranth emergence is
human-mediated by tillage timings and preemergence-applied herbicides
(Chahal et al., 2021), which can result in weed germination shifts
(Sbatella and Wilson, 2010). Also, Palmer amaranth has shown to grow and
develop under reduced soil water content. In continuous water stress,
Palmer amaranth survived and produced at least 14000 seeds
plant\textsuperscript{-1} (Chahal et al., 2018). Seeds from Palmer
amaranth growing with limited water conditions were heavier, less
dormant, and prompt for germination (Matzrafi et al., 2021). Growing
conditions and management practices also influence Palmer amaranth sex
dimorphism and flowering pattern (Korres et al., 2017; Rumpa et al.,
2019). Therefore, Palmer amaranth has shown plasticity to evolve and
fast adapt under the current agroecosystem conditions. Future scenarios
suggest that global temperature warming can impact agriculture, and
promote niches for Palmer amaranth invasion/adaptation into new
environments. Agronomic crops in Australia and Sub-Sahara Africa are
currently at greatest risk to Palmer invasion as climate warms (Kistner
and Hatfield, 2018). Temperature is a key factor limiting Palmer
amaranth expansion to cooler geographies (Briscoe Runquist et al.,
2019); however, under future climate change Palmer amaranth is likely to
expand northward into Canada and Northern Europe (Kistner and Hatfield,
2018; Briscoe Runquist et al., 2019).

Palmer amaranth is already found in agronomic crops of South America
(Larran et al., 2017; Küpper et al., 2017) and Southern Europe (Milani
et al., 2021). In the United States, Palmer amaranth is well established
in the Cotton Belt (Garetson et al., 2019; Bagavathiannan and
Norsworthy, 2016) in the southern United States but its range is
expanding northward. For example, herbicide resistant Palmer amaranth is
widespread in Nebraska (Oliveira et al., 2021). There are some reported
cases of Palmer amaranth in Michigan (Kohrt et al., 2017) and
Connecticut (Aulakh et al., 2021). Also, it is estimated that Palmer
amaranth can cause yield loss to soybean fields in Illinois (Davis et
al., 2015), which is concerning as soybean along with corn comprise most
of US Midwest agronomic hectares. In Iowa, a study showed that Palmer
amaranth is still not well adapted compared to waterhemp
(\emph{Amaranthus tuberculatus}) (Baker, 2021), which is considered the
most troublesome species in the US Midwest (Tranel et al., 2011).
Invasion and successful eradication of Palmer amaranth is documented in
Minnesota (Yu et al., 2021). Palmer amaranth infestations have not been
detected in Canada; however, Palmer amaranth seeds were detected in
sweet potato slips in the country (Page et al., 2021). Palmer amaranth
is still not as well adapted and established to Northern as compared to
Southern North America. Therefore, Palmer amaranth range of expansion
into new habitats can increase. It seems certain the need to manage new
Palmer amaranth infestations in agronomic crops throughout northern
United States is eminent in the near future. Strategies on Palmer
amaranth management should encompass the agroecosystem level but not
focus only on attempts to eradicate the weed. Most tactics to manage
Palmer amaranth are based on technological solutions (Scott, 2011),
which are short-term (e.g., herbicide and/or tillage) rather than
long-term weed management. Palmer amaranth management should be built
around minimizing the species ability to adapt, grow and develop
successfully in agroecossystems.

In the southeastern US, early growing Palmer amaranth is well known to
have a higher impact on cotton yields compared to late established
plants (MacRae et al., 2013). In the northern states, Palmer amaranth
impact on the agroecosystem is recent. Studies investigating Palmer
amaranth in northern states are limited due to the plant classification
as noxious weed species (Yu et al., 2021). Nonetheless, the continuous
Palmer amaranth dispersal and potential establishment across the
northern United States is concerning and warrants investigations on
species morphology in northern environments. Understanding Palmer
amaranth biology and morphology under different agroecosystems can
enhance our knowledge on species adaptation and management practices. It
can also aid in designing proactive and ecological tactics to limit the
species range expansion, reduce its negative impact on agroecossystems,
and to develop resilient and sustainable farming systems (MacLaren et
al., 2020). Therefore, the objective of this study was to investigate
the flowering pattern, gender, biomass production, and height of Palmer
amaranth cohorts growing under corn, soybean and bareground across five
locations in the United States Midwest.

@@

\caption{\label{tab:unnamed-chunk-2}Field study attributes from Arlington, WI, Clay Center, NE, Grant, NE, Lincoln, NE and Macomb, IL}
\centering
\resizebox{\linewidth}{!}{
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lllllll}
\toprule
Attributes &  & Arlington, WI & Clay Center, NE & Grant, NE & Lincoln, NE & Macomb, IL\\
\midrule
Bareground &  &  &  &  &  & \\

 & Weed control\textsuperscript{a} & \emph{S}-metolachor\textsuperscript{b} fb glyphosate\textsuperscript{c} /  & saflufenacil + imazethapyr + pyroxasulfone\textsuperscript{d} & glyphosate & glyphosate + \emph{S}-metolachor\textsuperscript{f} fb glyphosate\textsuperscript{g} & \\

@@

 & Hybrid & NK0142 3120-EZ1 & DKC60-67 & G03C84-5122-EZ1 & P1197AM & \\

 & Seeding rate & 88956 & 86487 & 79000 & 86485 & \\

 & Weed control & \emph{S}-metolachor fb glyphosate & \emph{S}-metolachlor + trazine + mesotrione + bicyclopyrone\textsuperscript{e} & glyphosate & glyphosate + \emph{S}-metolachor\textsuperscript{f} fb glyphosate\textsuperscript{g} & \\

 & Stage at 1\textsuperscript{st} cohort & V2-3 & V6-7 & V6-7 & V2-3 & \\

 & Stage at 2\textsuperscript{nd} cohort & V6-7 & V6-7 & V6-7 & V6-7 & \\

 & Planting day & April 30, 2018 / May 5, 2019 & May 10, 2018/19 & May 5, 2018 / May 14, 2019 & April 27, 2018 / May 3, 2019 & \\

\multirow{-7}{*}{\raggedright\arraybackslash } & Fertilization & N (46-0-0) at 157 kg ha\textsuperscript{-1} &  &  & N (46-0-0) at 336 kg ha\textsuperscript{-1} & \\

@@

 & Variety & DSR-1950 & AG21X8 & AG28X7 & 2018: 3017R2X; 2019 P31A22X & \\

 & Seeding rate & 296400 & 321237 & 220000 & 370650 & \\

 & Weed control & glyphosate / \emph{S}-metolachor & saflufenacil + imazethapyr + pyroxasulfone & glyphosate & glyphosate + \emph{S}-metolachor fb glyphosate & \\

 & Stage at 1\textsuperscript{st} cohort & V1-2 & V1-2 & V1-2 &  & \\

 & Stage at 2\textsuperscript{nd} cohort & V5-6 & V5-6 & V5-6 &  & \\

\multirow{-6}{*}{\raggedright\arraybackslash } & Planting day & May 5, 2018 / May 10, 2009 & May 14, 2018/19 & May 5, 2018 / May 14, 2019 & May 7, 2018 / May 17, 2019 & \\

@@

 & Type & Plano-silt-loam & Crete Silt Loam & 2018: Mace silt loam / 2019: Kuma silt loam & Crete silty clay loam & \\

 & pH & 6.6 & 6.5 & 2018: 6.1 / 2019: 5.1 & 5.2 & \\

\multirow{-4}{*}{\raggedright\arraybackslash } & Organic matter (\%) & 3.5 & 3 & 2018: 2 / 2019: 1.7 & 3.4 & \\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{a}Hoe weeding; \textsuperscript{b}\emph{S}--metolachor, 1324 g ai ha\textsuperscript{-1}; \textsuperscript{c}glyphoste, 1262 g ae ha\textsuperscript{-1}; \textsuperscript{d}saflufenacil + imazethapyr + pyroxasulfone, 215 g ai ha\textsuperscript{-1}; \textsuperscript{e}\emph{S}-metolachlor + trazine + mesotrione, + bicyclopyrone, 2409 g ai ha\textsuperscript{-1}; \textsuperscript{f}\emph{S}--metolachor, 1070 g ai ha\textsuperscript{-1}; \textsuperscript{g}glyphoste, 870 g ae ha\textsuperscript{-1}} \\
\end{tabular}}
\end{table}

The field experimental units were three adjacent 9.1 m wide (12 rows at
76.2 cm row spacing) by 10.7 m long. The experimental design were
arranged in factorial design with three crops, two transplanting times
simulating two cohorts, repeated across five locations. Each field
experimental unit was planted with corn, soybean, or kept under
bareground. The two transplant timings were June 1 (first cohort) and
July 1 (second cohort). Palmer amaranth seedlings (potting mix + two
seedlings) were transplanted (6 cm deep and 8 cm wide). Forty-eight
plants were equidistantly placed (0.76 m apart) between rows within each
crop. After a week, one plant was eliminated and one was kept, resulting
in 24 plants per experimental unit and transplanting time (Figure 2).
When needed, Palmer amaranth plants were supplied with water during the
first week after transplanting to assure seedling survival.

After transplanting, Palmer amaranth flowering was monitored until the
end of the study. When a plant flowered, the day was recorded, plant
gender was identified (male or female), plant height was measured from
soil surface to the top of plant. Also, aboveground plant biomass was
harvested near soil surface and oven dried at 65 C until reaching
constant weight before weigh (g plant\textsuperscript{-1}) was recorded.

@@

A binary logistic regression was fitted to Palmer amaranth gender.
Binary logistic regression is used for predicting binary classes
(Bangdiwala, 2018), such as the probability of a plant being female in a
dioecious species. Prior to the analysis, missing values were removed
from the dataset. Also, data from Grant was not used in this analysis
due to the uniform plant harvesting at that location. The complete
dataset was splitted into 80\% train and 20\% test data. The 80\% train
is used for the model training and the 20\% test is used for checking
model performance on unseen dataset. With 80\% dataset, a generalized
linear model (base R \emph{glm} function) was fitted to binary response
variable, the probability of being female (0 to male and 1 to female).
The independent variables were day of year harvest, height, weight, and
crop (without interaction). The model family was binomial with a logit
function. The model fit was assessed through pseudo R-squared values
(McFadden, Cox and Snell, Cragg and Uhler) and likelihood ratio using
\emph{nagelkerke} function from ``rcompanion'' package (Mangiafico,
2021). The marginal effects computation was performed with Average
Marginal Effects (AMEs) at every observed value of x and average across
the results (Leeper, 2017) using \emph{margins} function from
``margins'' package (Leeper et al., 2021). The 20\% test data was
predicted using the \emph{predict} function with a cutoff estimation for
male or female using \emph{performance} function from ROCR package (Sing
et al., 2005). The model quality prediction from the classification
algorithm was measured with precision (\emph{precision} function),
recall (\emph{recall} function) and F1-score (\emph{f\_meas} function)
using the ``yardstick'' package (Kuhn et al., 2021). The precision
determines the accuracy of positive predictions (female plants), recall
determines the fraction of positives that were correctly identified, and
F1-score is a weighted harmonic mean of precision and recall with the
best score of 1 and the worst score of 0 (Raoniar, 2021). F1-score
conveys the balance between the precision and the recall (Yacouby and
Axman, 2020). The area under the receiver operating curve (AUC-ROC) was
also estimated with performance function using the true positive and
false positive rates. Higher the AUC, better the model is at
distinguishing between female and male Palmer amaranth.

@@

\hypertarget{palmer-amaranth-gender}{%
\subsection*{Palmer amaranth gender}\label{palmer-amaranth-gender}}
\addcontentsline{toc}{subsection}{Palmer amaranth gender}

@@

\caption{\label{tab:unnamed-chunk-3}Average marginal effects of Palmer amaranth gender logistic model. Factor pararemter values (crop and bareground) is shown compared to soybean.}
\centering
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{lrrrrrr}
\toprule
Term & AME & SE & Lower & Upper & Z-score & P-value\\
\midrule
crop\_bareground & -0.048 & 0.054 & -0.154 & 0.059 & -0.876 & 0.381\\

@@

Our study showed that Palmer amaranth biomass, height, flowering window
and gender varied within crops and cohort timings. In general, first
cohort of Palmer amaranth plants were heavier and taller when compared
to the second cohort. At first cohort, resources (e.g., soil nutrients)
and conditions (e.g., light) were more timely available for the species.
High biomass and taller Palmer amaranth plants are likely a weed
strategy to compete for light in between crop rows in absence of canopy.
In such conditions, Palmer amaranth showed an extraordinary plasticity
to adapt upon the agroecosystem. This is evident when comparing Palmer
amaranth canopy shape. The Palmer amaranth competition (e.g., light)
strategy was to mimic the crop growth and development (Figure 6). These
results suggests that Palmer amaranth can quickly evolve life-history
traits to adapt to cropping systems and cultural practices, which was
also shown in a study varying nitrogen fertilization (Bravo et al.,
2018). Our results highlight Palmer amaranth as a threat to field crops
as breeding more competitive crop varieties is likely to select more
competitive weed biotypes (Bravo et al., 2017).

Palmer amaranth growth and development in second cohort was limited due
to the crop competitive ability at advanced development stages. Plants
were transplanted at greater crop height and width, which reduced Palmer
amaranth competitiveness. As a result, Palmer amaranth height and
biomass was lower compared to its first cohort. Moreover, Palmer
amaranth growing without crop competition produced the highest amounts
of biomass. The Palmer amaranth strategy in bareground was to invest
biomass in growing plant width and height. Nonetheless, Palmer amaranth
produced 17\% less biomass in second cohort compared to first cohort
timing. In a bareground study, early emerged Palmer amaranth without
competition was 50\% taller than late emerged plants (Webster and Grey,
2015). These results suggest that crop competition is not the only
factor limiting late Palmer amaranth establishment. The limited growth
of Palmer amaranth at second cohort is likely a reduced plant response
to day length, light availability and thermal units (e.g, growing degree
days). The \emph{Amaranthus} species are sensitive to photoperiod (Wu
and Owen, 2014). It is hypothesize that reduced day length or
red/far-red effect contributed to smaller plants at second cohort
regardless the crop. A study in North Carolina and Illinois predicted
that less than 10\% Palmer amaranth seedlings emergence occurred after
June (Piskackova et al., 2021). In addition, Palmer amaranth negative
impact on soybean (Korres et al., 2020) and cotton (Webster and Grey,
2015) yields was higher when plants were established near to crop
planting. Therefore, early management is a key strategy to minimize the
damaging impact of Palmer amaranth to US Midwest cropping systems.

Seed production was not evaluated due to plant harvest at initiation of
flowering. Nonetheless, it is well documented a strong positive
correlation between Palmer amaranth biomass and seed production
(Schwartz et al., 2016; Spaunhorst et al., 2018). In our study, plants
growing from the first cohort accumulated 42\% more biomass when
compared to the second cohort. Therefore, Palmer amaranth plants growing
in the second cohort is likely to produce less seeds regardless the
crop. Our observation is consistent with the findings that the first
Palmer amaranth cohort produced 50\% more seeds per plant than Palmer
amaranth plants established six weeks later in bareground (Webster and
Grey, 2015). Still, seed production at the second cohort will replenish
the soil seedbank. Seed production and deposition in the seedbank is
also a key factor for species perpetuation (Menges, 1987). Palmer
amaranth can produce a hundred thousands seeds per plant (Schwartz et
al., 2016; Keeley et al., 1987), which can stay viable in the soil
seedbank for at least 36 months (Sosnoskie et al., 2013). Therefore,
preventing Palmer amaranth seed production or/and seed migration to
non-native habitats is an essential strategy to minimize the species
impact in agroecosystems (Davis et al., 2015).
